### 2018-01-25 submission/manuscript.Rmd
added first two sections of results ada suggestions

**Results.** Fecal samples had a significant decrease from those without disease to adenoma (less severe disease) to carcinoma (severe disease) for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
Power transforming and Z-score normalizing Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether these detectable differences in community resulted in significant increases in RR of lesion (adenoma or carcinoma). For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].

Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. For tissue, there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C]. When analyzing the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) tissue samples, for both adenoma and carcinoma there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5]. Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were also identified in unmatched tissue samples  for both those with adenoma and carcinoma [Table S6 & S7]. For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].

***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
Based on the small increase in RR using OTU richness and Shannon diversity, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue. To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3]. Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3]. Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.

For adenoma, there was almost no overlap between genera with a RR P-value of less than 0.05 and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had a P-value under 0.05 for either an increased or dereased RR are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both tissue and stool [Table S10 & S11]. Although the most significant increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

>>>>

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness after correcting for study effect and variable region sequenced with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples. A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples. Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized. These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma. For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].

Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased. For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C]. We next analyzed the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples. For individuals at either an adenoma or carcinoma stage of disease there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5]. Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S6 & S7]. For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7]. These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.

***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
Next we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. The top 5 most significantly increased RR were the same between adenoma and carcinoma for both stool and colon tissue [Figure 3]. Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of CRC, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

### 2018-01-25 submission/manuscript.tex
added first two sections of results ada suggestions

\textbf{Results.} Fecal samples had a significant decrease from those
without disease to adenoma (less severe disease) to carcinoma (severe
disease) for both Shannon diversity and evenness after correcting for
study effect and variable region sequenced (P-value \textless{} 0.05).
This reduction in evenness translated into small increases in relative
risk for adenoma (P-value = 0.032) and carcinoma stages of CRC (P-value
= 0.00034) while the reduction in Shannon diversity only translated into
an increased relative risk for developing carcinomas (P-value = 0.0047).
Increases in mouth-associated microbes were commonly in the top 5 most
significantly increased relative risk of adenoma and carcinoma for both
stool and tissue samples. A prediction model for adenoma and carcinoma
was built using either the whole community or selected genera with
highest and lowest relative risk from fecal and tissue samples. Both
approaches resulted in similar classification success according to Area
Under the Curve (AUC) regardless of whether genera or OTUs were used to
build the model. The most important groups within the full community
models consistently belonged to genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
number of associations between the microbiota and CRC were identified,
the majority of studies that we used in this meta-analysis were only
individually adequately powered for large effect sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Carcinomas:}} Power transforming and Z-score normalizing Operational
Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the
entire data set allowed for the combination of data from all studies to
assess differences in broad scale community metrics as disease severity
worsens. Using linear mixed-effect models to control for study and
variable region we assessed whether OTU richness, evenness, or Shannon
diversity changed in a step-wise manner with disease severity. In stool,
there was a significant decrease in both evenness and Shannon diversity
from control to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1A{]}. We next tested whether these detectable
differences in community resulted in significant increases in RR of
lesion (adenoma or carcinoma). For fecal samples, a decrease versus the
overal median in evenness resulted in a significantly increased RR for
carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034) and adenoma (RR =
1.16 (1.01 - 1.34), P-value = 0.032) while a decrease versus the overal
median in Shannon diversity only increased the RR for carcinoma (RR =
1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Using the
Bray-Curtis distance metric and PERMANOVA, it was also possible to
identify significant bacterial community changes, in specific studies,
for both carcinoma versus control and adenoma versus control {[}Table S1
\& S2{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and variable region to test whether
OTU richness, evenness, or Shannon diversity changed in a step-wise
manner as disease severity increased. For tissue, there were no
significant changes in OTU richness, evenness, and Shannon diversity
from control to adenoma to carcinoma (P-value \textgreater{} 0.05)
{[}Figure 1B \& C{]}. When analyzing the RR, for matched (unaffected
tissue and an adenoma or carcinoma from the same individual) and
unmatched (control and adenoma or carcinoma tissue not from the same
individual) tissue samples, for both adenoma and carcinoma there was no
significant change in RR based on lower than median values for OTU
richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar to
stool samples, significant differences in bacterial community, assessed
by PERMANOVA, were also identified in unmatched tissue samples for both
those with adenoma and carcinoma {[}Table S6 \& S7{]}. For studies with
matched samples no differences in bacterial community were observed when
assessed with PERMANOVA {[}Table S6 \& S7{]}.

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of Lesion:}} Based on the small increase in RR using OTU richness and
Shannon diversity, we assessed if being higher than the median for any
specific genera resulted in an altered RR for adenoma or carcinoma in
stool and tissue. To investigate this we analyzed all common genera
across each study, in tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. For both stool and tissue none of the genera in the top 5 most
significantly increased RR were the same between adenoma and carcinoma
{[}Figure 3{]}. Mouth-associated genera were in the top 5 genera
associated with an increased RR of adenoma (\emph{Pyramidobacter}
{[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]}) and carcinoma
(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas}, and
\emph{Peptostreptococcus} {[}Figure 3B{]} and \emph{Fusobacterium}
{[}Figure 3D{]}) for both stool and tissue. Conversely, genera commonly
associated with a normal gastrointestinal tract were correlated with a
decreased RR for both adenoma and carcinoma for both stool and tissue
{[}Figure 3{]}. Overall, there was little direct overlap of the top 5
increased or decreased RR genera between stool and tissue.

For adenoma, there was almost no overlap between genera with a RR
P-value of less than 0.05 and when they were similar the RR was in
opposite directions (e.g. \emph{Lactococcus}) {[}Table S8 \& S9{]}. Many
of the genera that had a P-value under 0.05 for either an increased or
dereased RR are also highly prevalent in contamination, specifically,
\emph{Novosphingobium}, \emph{Selemonas}, and \emph{Achromobacter}
{[}Figure 3 \& Table S8-S9{]}. For carcinoma, certain mouth-associated
genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}) had a high RR for both tissue and stool
{[}Table S10 \& S11{]}. Although the most significant increased RR
genera for tissue was \emph{Camplyobacter} while in stool it was
\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.

>>>>

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness after correcting for study effect and
variable region sequenced with more severe disease (P-value \textless{}
0.05). This reduction in evenness translated into small increases in
relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
(P-value = 0.00034) while the reduction in Shannon diversity only
translated into an increased relative risk for developing carcinomas
(P-value = 0.0047). Increases in mouth-associated microbes were commonly
in the top 5 most significantly increased relative risk of adenoma and
carcinoma for both stool and tissue samples. A prediction model for
adenoma and carcinoma was built using either the whole community or
selected genera with highest and lowest relative risk from fecal and
tissue samples. Both approaches resulted in similar classification
success according to Area Under the Curve (AUC) regardless of whether
genera or OTUs were used to build the model. The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
CRC were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of Carcinomas:}} To assess differences in broad scale community metrics
as disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overal median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overal median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

Using similar transformations for tissue samples, linear mixed-effect
models were used on the transformed combined data to control for study,
re-sampling of the same individual, and 16S variable region to test
whether OTU richness, evenness, or Shannon diversity changed in a
step-wise manner as disease severity increased. For colon tissue, there
were no significant changes in OTU richness, evenness, or Shannon
diversity as disease severity progressed from control to adenoma to
carcinoma (P-value \textgreater{} 0.05) {[}Figure 1B \& C{]}. We next
analyzed the RR, for matched (unaffected tissue and an adenoma or
carcinoma from the same individual) and unmatched (control and adenoma
or carcinoma tissue not from the same individual) colon tissue samples.
For individuals at either an adenoma or carcinoma stage of disease there
was no significant change in RR based on lower than median values for
OTU richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar
to stool samples, significant differences in bacterial community,
assessed by PERMANOVA, were identified in unmatched tissue samples, for
those at either adenoma or carcinoma stage of CRC {[}Table S6 \& S7{]}.
For studies with matched samples no differences in bacterial community
were observed when assessed with PERMANOVA {[}Table S6 \& S7{]}. These
tissue results suggest that the microbiota within an individual are
similar to each other regardless of disease status.

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of CRC:}} Next we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered RR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in RR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. The top 5 most significantly increased RR were the same between
adenoma and carcinoma for both stool and colon tissue {[}Figure 3{]}.
Mouth-associated genera were commonly found in the top 5 genera
associated with an increased RR of having an adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased RR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased RR genera between both stages and sample site.

When observing RRs with a P-value less than 0.05 there was almost no
overlap between genera from stool or colon tissue and when they were
similar the RR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the genera that had RRs with a P-value
under 0.05 for colon tissue are also highly prevalent in contamination,
specifically, \emph{Novosphingobium}, \emph{Selemonas}, and
\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
of CRC, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high RR for both
colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
data suggest that the most significantly increased RR genera for tissue
was \emph{Camplyobacter} while in stool it was \emph{Peptostreptococcus}
{[}Table S10 \& S11{]}.
